SOURCE SEPARATION METHODS FOR COMPUTER-ASSISTED ORCHESTRATION

### 2021-06-02 source_separation/paper/paper.tex
Update file to use MLSP paper style

% -----------------------------------------------
% Template for SMC 2021
% Adapted from previous SMC paper templates
% -----------------------------------------------
\documentclass{article}
\usepackage{smc2021}
%%%%%%%%%%%%%%%%%%%%%%%% Some useful packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%% See related documentation %%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[caption=false, font=footnotesize]{subfig}% Modern replacement for subfigure package
\usepackage{paralist}% extended list environments
\usepackage[figure,table]{hypcap}% hyperref companion
\usepackage{multirow}
% Enable for Review only, remove for Camera Ready version
\pagewiselinenumbers

% Use this if english is the only language/alphabet used in the document
\usepackage[english]{babel}

% Title.
% ------
\def\papertitle{Source separation methods for computer-assisted orchestration}

% Authors
% Please note that submissions are NOT anonymous, therefore 
% authors' names have to be VISIBLE in your manuscript. 
% Authors are entered as an ordered list, each one can be linked to multiple affiliations using the correct index.
% Available tags for authors are: \firstname \middlename \lastname \generation \originalname \email \orcid
% Available tags for affiliations are: \unit \department \institution \streetaddress \city \state \postcode \country \type
% type can take as value: University, Company, Music, Independent, Other
%
% \author[]{\mbox{\firstname{}\middlename{}\lastname{}\originalname{}\generation{}\email{}\orcid{}}}
% mbox force an author not to be split over multiple lines
\author[1]{\mbox{\firstname{Luke}\lastname{Dzwonczyk}\email{dz.luke@berkeley.edu}}}
\author[1, 2, 3]{\mbox{\firstname{L\'eo}\lastname{Ch\'edin}}}
\author[1]{\mbox{\firstname{Alejandro}\lastname{Saldarriaga-Fuertes}}}
\author[1]{\mbox{\firstname{Max}\lastname{Sherr}}}
\author[3]{\mbox{\firstname{H\'el\`ene-Camille}\lastname{Crayencour}}}
\author[1]{\mbox{\firstname{Carmine-Emanuele}\lastname{Cella}}}

%%Affiliations
\affil[1]{\department{Center for New Music and Audio Technologies}\institution{University of California, Berkeley}\city{Berkeley}\state{California}\country{USA}\affiliationtype{University}}
\affil[2]{\department{\'Ecole Normale Sup\'erieure Paris-Saclay}\institution{Universit\'e Paris-Saclay}\city{Paris}\state{}\country{France}\affiliationtype{University}}
\affil[3]{\department{Laboratoire des Signaux et Syst\`emes}\institution{Centrale Sup\'elec, CNRS, Universit\'e Paris-Saclay}\city{Paris}\state{}\country{France}\affiliationtype{University}}

% Complete setup stage
\completesetup

% Title.
% ------
\title{\papertitle}
% ***************************************** the document starts here ***************
\begin{document}
	%
	\capstartfalse
	\maketitle
	\capstarttrue
	%

	\begin{abstract}
		In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
	\end{abstract}
	%

	\section{Introduction}\label{sec:introduction}

	Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes. In particular, target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints~\cite{Maresz2003}.

@@

	The codebase for this paper can be found at: \mbox{\url{https://github.com/dzluke/smc2021}}, and you can listen to a selection of targets and orchestrations here: \mbox{\url{https://dzluke.github.io/smc2021/}}.

@@

	We test the effectiveness of five difference source separation methods, both supervised and unsupervised, on 300 custom made targets that are combinations of sounds from different databases. During testing, a target is input to a source separation method, which outputs four sub-targets. Then each sub-target is independently orchestrated with a randomly assigned subset of the full orchestra. These orchestrations are then combined to play simultaneously, creating a final orchestrated solution. Then the distance between the target and solution is calculated, giving us a metric to compare the various separation methods. The random split of the orchestra is the same for each separation method of the same target. See Fig. \ref{fig:eval_diagram} for a diagram of this process.

		\subsection{Data}\label{subsec:data}
		We created our own targets as combinations of four source sounds. The sources come from the NIGENS \cite{NIGENS} and BBC \cite{BBC} databases, and freesound.org \cite{freesound}. We selected a total of 90 samples from these databases, choosing sounds that fit the following criteria:

@@

		As described in Section \ref{subsec:data}, each target is comprised of four sources that are offset in time to begin playing at different times. Figure \ref{fig:orchestrations} illustrates the effect this offset has on the full target, separated, and ground truth orchestrations. In the full target orchestration, all onsets occur at the beginning of the orchestration. Since there is no knowledge of different sources, Orchidea cannot place the different orchestrations at the appropriate offsets. For the separated orchestrations, the onsets of the sub-target orchestrations can be placed at the correct offset if the separation method has correctly separated out the sources. In the ground truth orchestration, the sources are manually offset, leading to the ground truth having each source orchestration beginning at the correct time. For this reason, despite only performing static orchestrations, the resulting separated and ground truth orchestrations are offset in time compared to the full target orchestration.

@@

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	%bibliography here
	\bibliography{references}

>>>>

% LaTeX template for MLSP papers. To be used with:
%   * mlspconf.sty - ICASSP/ICIP LaTeX style file adapted for MLSP, and
%   * IEEEbib.bst - IEEE bibliography style file.
% --------------------------------------------------------------------------
\documentclass{article}
\usepackage{amsmath,graphicx,mlspconf}
%%%%%%%%%%%%%%%%%%%%%%%% Some useful packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%% See related documentation %%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[caption=false, font=footnotesize]{subfig}% Modern replacement for subfigure package
\usepackage{paralist}% extended list environments
\usepackage{url}
\usepackage{hyperref}
\usepackage[figure,table]{hypcap}% hyperref companion
\usepackage{multirow}

\usepackage{amssymb}

% Copyright notices.
% ------------------
% Select one of the four copyright notices below. Only required for the camera-ready paper submission.
% 
% * For papers in which all authors are employed by the US government:
\copyrightnotice{U.S.\ Government work not protected by U.S.\ copyright}

% * For papers in which all authors are employed by a Crown government (UK, Canada, and Australia):
\copyrightnotice{978-1-7281-6338-3/21/\$31.00 {\copyright}2021 Crown}

% * For papers in which all authors are employed by the European Union:
\copyrightnotice{978-1-7281-6338-3/21/\$31.00
{\copyright}2021 European Union}

% * For all other papers:
\copyrightnotice{978-1-7281-6338-3/21/\$31.00 {\copyright}2021 IEEE}

% Header
\toappear{2021 IEEE International Workshop on Machine Learning for Signal Processing, Oct.\ 25--28, 2021, Gold Coast, Australia}

% Example definitions.
% --------------------
%\def\x{{\mathbf x}}
%\def\L{{\cal L}}

% Title.
% ------
\title{SOURCE SEPARATION METHODS FOR COMPUTER-ASSISTED ORCHESTRATION}
%
% Double-blind peer review.
% -------------------------
% Anonymize your paper for the double-blind peer-review process using the 
% following author and affiliation.
\name{Anonymous\thanks{Anonymous.}}
\address{Anonymous}

% Single address.
% ---------------
%\name{Author(s) Name(s)\thanks{Thanks to XYZ agency for funding.}}
%\address{Author Affiliation(s)}

% For example:
% ------------
%\address{%
%    School \\
%    Department \\
%    Address
%}
%
% Two addresses.
% --------------
%\twoauthors{%
%    A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}
%}{%
%    School A-B \\
%    Department A-B \\
%    Address A-B \\
%    Email A-B
%}{%
%   C. Author-three, D. Author-four\sthanks{The fourth author performed the work while at ...}
%}{%
%    School C-D \\
%    Department C-D \\
%    Address C-D \\
%    Email C-D
%}
% 
% Two or more addresses (alternative form).
% -----------------------------------------
% If you need to list more than 2 authors or the option for two options above 
% produces a poor author block, please use the following structure:
%\name{%
%    Author Name$^{\star \dagger}$%
%    \qquad Author Name$^{\star}$%
%    \qquad Author Name$^{\dagger}$\thanks{Thanks to XYZ agency for funding.}%
%}
%\address{%
%    $^{\star}$ Affiliation Number One \\%
%    $^{\dagger}$ Affiliation Number Two%
%}

\begin{document}
%\ninept

\maketitle

\begin{abstract}
  In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
\end{abstract}
%
\begin{keywords}
One, two, three, four, five
\end{keywords}
%
\section{Introduction}\label{sec:introduction}
  Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes. In particular, target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints~\cite{Maresz2003}.

@@

%  The codebase for this paper can be found at: \url{https://github.com/dzluke/smc2021}, and you can listen to a selection of targets and orchestrations here: \url{https://dzluke.github.io/smc2021/}.
  The codebase for this paper can be found at: \url{anonymous}, and you can listen to a selection of targets and orchestrations here: \url{anonymous}.

@@

  We test the effectiveness of five difference source separation methods, both supervised and unsupervised, on 300 custom made targets that are combinations of sounds from different databases. During testing, a target is input to a source separation method, which outputs four sub-targets. Then each sub-target is independently orchestrated with a randomly assigned subset of the full orchestra. These orchestrations are then combined to play simultaneously, creating a final orchestrated solution. Then the distance between the target and solution is calculated, giving us a metric to compare the various separation methods. The random split of the orchestra is the same for each separation method of the same target. See Fig.~\ref{fig:eval_diagram} for a diagram of this process.

    \subsection{Data}\label{subsec:data}
    We created our own targets as combinations of four source sounds. The sources come from the NIGENS~\cite{NIGENS} and BBC~\cite{BBC} databases, and freesound.org \cite{freesound}. We selected a total of 90 samples from these databases, choosing sounds that fit the following criteria:

@@

    As described in Section \ref{subsec:data}, each target is comprised of four sources that are offset in time to begin playing at different times. Figure~\ref{fig:orchestrations} illustrates the effect this offset has on the full target, separated, and ground truth orchestrations. In the full target orchestration, all onsets occur at the beginning of the orchestration. Since there is no knowledge of different sources, Orchidea cannot place the different orchestrations at the appropriate offsets. For the separated orchestrations, the onsets of the sub-target orchestrations can be placed at the correct offset if the separation method has correctly separated out the sources. In the ground truth orchestration, the sources are manually offset, leading to the ground truth having each source orchestration beginning at the correct time. For this reason, despite only performing static orchestrations, the resulting separated and ground truth orchestrations are offset in time compared to the full target orchestration.

@@

% References should be produced using the bibtex program from suitable
% BiBTeX files (here: strings, refs, manuals). The IEEEbib.bst bibliography
% style file from IEEE produces unsorted bibliography list.
% -------------------------------------------------------------------------
\bibliographystyle{IEEEbib}
\bibliography{refs}
